---
title: "DSP 110 Project - Comic Character Analysis"
output:
  html_notebook: default
  pdf_document: default
---

# Setup and Data Loading

```{r setup, message=FALSE, warning=FALSE}
library(readr)
library(dplyr)
library(tidyr)
library(lubridate)
library(janitor)
library(coursekata)
library(curl)
```

## Load Data from GitHub

```{r load-data}
# Get DC characters:
comic_characters_dc <- 
  "https://github.com/fivethirtyeight/data/raw/master/comic-characters/dc-wikia-data.csv" %>% 
  read_csv() %>% 
  clean_names() %>% 
  mutate(publisher = "DC")

# Get Marvel characters:
comic_characters_marvel <- 
  "https://github.com/fivethirtyeight/data/raw/master/comic-characters/marvel-wikia-data.csv" %>% 
  read_csv() %>% 
  clean_names() %>% 
  mutate(publisher = "Marvel")
```

## Data Wrangling

```{r wrangle-data}
# Merge two datasets and perform further data wrangling:
comic_characters <-
  comic_characters_dc %>% 
  bind_rows(comic_characters_marvel) %>% 
  separate(first_appearance, c("year2", "month"), ", ", remove = FALSE) %>%
  mutate(
    # If month was missing, set as January and day as 01:
    month = ifelse(is.na(month), "01", month),
    day = "01",
    # Note some years missing:
    date = ymd(paste(year, month, day, sep = "-")),
    align = factor(
      align, 
      levels = c("Bad Characters", "Reformed Criminals", "Netural Characters", "Good Characters"),
      ordered = TRUE)
  ) %>%
  select(publisher, everything(), -c(year2, day))
```

## Data Exploration

```{r explore-structure}
str(comic_characters)
```

```{r explore-head}
head(comic_characters)
```

```{r explore-sex}
unique(comic_characters$sex)
```

```{r explore-id}
unique(comic_characters$id)
```

## Data Cleaning

```{r clean-data}
comic_clean <- comic_characters %>%
    mutate(sex = recode(sex,
                      "Male Characters" = "Male",
                      "Female Characters" = "Female",
                      "Agender Characters" = "Agender",
                      "Genderfluid Characters" = "Genderfluid",
                      "Genderless Characters" = "Genderless")) %>%
    filter(!is.na(sex), !is.na(id))
```

```{r exclusion-summary}
# Calculate exclusion summary
total_rows <- nrow(comic_characters)
cleaned_rows <- nrow(comic_clean)
excluded_rows <- total_rows - cleaned_rows
excluded_percent <- (excluded_rows / total_rows) * 100

cat("Rows removed due to missing sex or identity:", excluded_rows, 
    "(", round(excluded_percent, 2), "% of total )\n")
```

# Research Question 1: Alive Status by Publisher

```{r alive-by-publisher}
valid_alive <- c("Living Characters", "Deceased Characters")

# DC Characters
barplot(
  table(comic_characters_dc$alive[comic_characters_dc$alive %in% valid_alive]),
  main = "DC Characters Alive vs Dead",
  xlab = "Alive Status",
  ylab = "Count",
  col = c("red", "blue")
)

# Marvel Characters
barplot(
  table(comic_characters_marvel$alive[comic_characters_marvel$alive %in% valid_alive]),
  main = "Marvel Characters Alive vs Dead",
  xlab = "Alive Status",
  ylab = "Count",
  col = c("red", "blue")
)
```

# Research Question 2: Secret Identity by Gender

## Overall Analysis

```{r identity-by-gender}
by_gender <- comic_clean %>%
  count(sex, id) %>%
  group_by(sex) %>%
  mutate(prop_within_gender = n / sum(n)) %>%
  arrange(sex, desc(prop_within_gender))

by_gender
```

```{r overall-identity}
overall_id <- comic_clean %>%
  count(id) %>%
  mutate(prop_overall = n / sum(n)) %>%
  arrange(desc(prop_overall))

overall_id
```

## Male vs Female Focus

```{r gender-mf-analysis}
by_gender_MF <- comic_clean %>%
  filter(sex %in% c("Male", "Female")) %>%
  count(sex, id) %>%
  group_by(sex) %>%
  mutate(prop_within_gender = n / sum(n)) %>%
  arrange(sex, desc(prop_within_gender))

by_gender_MF
```

## Visualizations

```{r gender-identity-plot, fig.width=8, fig.height=6}
gf_bar(~ sex, fill = ~ id, data = comic_clean, position = "fill") %>%
  gf_labs(
    title = "Do Genders Differ in Having a Secret Identity?",
    y = "Proportion",
    x = "Gender"
  ) %>%
  gf_theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

```{r gender-counts-plots, fig.width=10, fig.height=5}
# Counts plot
gf_bar(~ sex, data = comic_clean) %>%
  gf_labs(title = "Number of Characters by Gender",
          y = "Count", x = "Gender")

# Proportions plot
comic_clean %>% 
  filter(id %in% c("Secret Identity", "Public Identity")) %>%
  gf_bar(~ sex, fill = ~ id, position = "fill") %>%
  gf_labs(title = "Secret vs Public Identities by Gender",
          y = "Proportion", x = "Gender") %>%
  gf_theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

## Probability Calculations

```{r prob-calculations}
prob_table <- comic_clean %>%
  count(sex, id) %>%
  group_by(sex) %>%
  mutate(prop = n / sum(n)) %>%
  filter(id == "Secret Identity")

prob_table %>%
  arrange(desc(prop))

male_secret <- prob_table$prop[prob_table$sex == "Male"]
female_secret <- prob_table$prop[prob_table$sex == "Female"]
diff <- male_secret - female_secret

cat("P(Secret | Male) =", round(male_secret, 4), "\n")
cat("P(Secret | Female) =", round(female_secret, 4), "\n")
cat("Difference =", round(diff, 4), "\n")
```

## DC-Specific Analysis

```{r dc-gender-analysis}
# Filter for DC characters only
dc_data <- comic_clean %>%
  filter(publisher == "DC", id %in% c("Secret Identity", "Public Identity"))

p_secret_male_dc <- mean(dc_data$id[dc_data$sex == "Male"] == "Secret Identity", na.rm = TRUE)
p_secret_female_dc <- mean(dc_data$id[dc_data$sex == "Female"] == "Secret Identity", na.rm = TRUE)
diff_dc <- p_secret_male_dc - p_secret_female_dc

cat("DC Comics:\n")
cat("P(Secret | Male) =", round(p_secret_male_dc, 4), "\n")
cat("P(Secret | Female) =", round(p_secret_female_dc, 4), "\n")
cat("Difference =", round(diff_dc, 4), "\n")
```

## Female Characters Over Time

```{r female-characters-timeline, fig.width=10, fig.height=6}
# Filter for female characters with valid year and identity data
female_data <- comic_clean %>%
  filter(sex == "Female", 
         !is.na(year)) %>%
  mutate(year = as.numeric(year))

# Create scatterplot with jitter
gf_jitter(id ~ year, data = female_data, color = ~ id, 
          height = 0.2, alpha = 0.6, size = 2) %>%
  gf_labs(
    title = "Female Characters' Identity Type by Year of Introduction",
    x = "Year Introduced",
    y = "Identity Type",
    color = "Identity Type"
  ) %>%
  gf_theme(theme_minimal()) %>%
  gf_theme(legend.position = "bottom")

# Summary statistics by decade
female_decade <- female_data %>%
  mutate(decade = floor(year / 10) * 10) %>%
  group_by(decade, id) %>%
  summarise(count = n(), .groups = "drop") %>%
  group_by(decade) %>%
  mutate(prop = count / sum(count))

# Plot proportion over time
gf_col(prop ~ decade, fill = ~ id, data = female_decade, position = "fill") %>%
  gf_labs(
    title = "Proportion of Female Characters with Secret vs Public Identity by Decade",
    x = "Decade",
    y = "Proportion",
    fill = "Identity Type"
  ) %>%
  gf_theme(theme_minimal()) %>%
  gf_refine(scale_y_continuous(labels = scales::percent))

# Count female characters introduced each year
female_year_counts <- female_data %>%
  group_by(year) %>%
  summarise(count = n(), .groups = "drop")

# Plot number introduced per year
gf_col(count ~ year, data = female_year_counts) %>%
  gf_labs(
    title = "Number of Female Characters Introduced Each Year",
    x = "Year",
    y = "Count of Female Characters"
  ) %>%
  gf_theme(theme_minimal())

male_data <- comic_clean %>%
  filter(sex == "Male", 
         !is.na(year)) %>%
  mutate(year = as.numeric(year))

male_year_counts <- male_data %>%
  group_by(year) %>%
  summarise(count = n(), .groups = "drop")

gf_col(count ~ year, data = male_year_counts) %>%
  gf_labs(
    title = "Number of Male Characters Introduced Each Year",
    x = "Year",
    y = "Count of Female Characters"
  ) %>%
  gf_theme(theme_minimal())

# Calculate proportions for modeling
female_prop <- female_data %>%
  group_by(year) %>%
  summarise(
    total = n(),
    secret_count = sum(id == "Secret Identity"),
    prop_secret = secret_count / total
  )

empty_model <- lm(prop_secret ~ NULL, data = female_prop)

cat("Empty Model Summary:\n")
summary(empty_model)
cat("\nBaseline (mean proportion of secret identities):", round(coef(empty_model), 4), "\n")

# Model with year as predictor
year_model <- lm(prop_secret ~ year, data = female_prop)

cat("\n\nYear Model Summary:\n")
summary(year_model)

# Compare models
cat("\n\nModel Comparison:\n")
cat("Empty Model R-squared:", round(summary(empty_model)$r.squared, 4), "\n")
cat("Year Model R-squared:", round(summary(year_model)$r.squared, 4), "\n")

# Scatterplot with regression line
gf_point(prop_secret ~ year, data = female_prop, 
         size = ~ total, alpha = 0.6, color = "purple") %>%
  gf_lm(color = "blue", linewidth = 1.5) %>%
  gf_labs(
    title = "Female Characters: Proportion with Secret Identity Over Time",
    subtitle = "With linear regression trendline and 95% confidence interval",
    x = "Year Introduced",
    y = "Proportion with Secret Identity",
    size = "# Characters"
  ) %>%
  gf_theme(theme_minimal()) %>%
  gf_refine(scale_y_continuous(labels = scales::percent))
# Filter for male characters

male_data <- comic_clean %>%
filter(sex == "Male", !is.na(year)) %>%
mutate(year = as.numeric(year))

# Calculate proportion secret by year for each sex

male_prop <- male_data %>%
group_by(year) %>%
summarise(
total = n(),
secret_count = sum(id == "Secret Identity"),
prop_secret = secret_count / total,
sex = "Male",
.groups = "drop"
)

female_prop <- female_data %>%
group_by(year) %>%
summarise(
total = n(),
secret_count = sum(id == "Secret Identity"),
prop_secret = secret_count / total,
sex = "Female",
.groups = "drop"
)

combined_prop <- bind_rows(female_prop, male_prop)

# Scatterplot with trendlines for both male & female

gf_point(prop_secret ~ year,
data = combined_prop,
color = ~sex,
size = ~total,
alpha = 0.6) %>%
gf_lm(color = ~sex, linewidth = 1.4) %>%
gf_labs(
title = "Proportion with Secret Identity Over Time (Male vs Female)",
subtitle = "Each sex shown with its own trendline and weighted point sizes",
x = "Year Introduced",
y = "Proportion with Secret Identity",
color = "Sex",
size = "# Characters"
) %>%
gf_theme(theme_minimal()) %>%
gf_refine(scale_y_continuous(labels = scales::percent)) %>%
gf_theme(legend.position = "bottom")

```

# Research Question 3: Identity Type Distribution

```{r identity-type-distribution}
valid_ids <- c("Secret Identity", "Public Identity")

# DC Characters
barplot(
  table(comic_characters_dc$id[comic_characters_dc$id %in% valid_ids]),
  main = "DC Characters by Identity Type",
  xlab = "Identity Type",
  ylab = "Count"
)

# Marvel Characters
barplot(
  table(comic_characters_marvel$id[comic_characters_marvel$id %in% valid_ids]),
  main = "Marvel Characters by Identity Type",
  xlab = "Identity Type",
  ylab = "Count"
)
```

# Conclusions

We wanted to know if a character's sex/gender affects their likelihood of having a secret identity, whether P(Secret | Male) is approximately equal to P(Secret | Female). If those two probabilities are roughly equal then gender and identity status can be considered independent.

Male and female characters dominate the dataset when it comes to the sex variable. There is not enough data to make a meaningful conclusion about agender, genderfluid, or genderless characters. There is a notable imbalance of about a 3:1 ratio of male to female characters in this dataset, but their proportion of secret identities are almost identical.

Since P(Secret | Male) = 0.499 and P(Secret | Female) = 0.480, the difference in proportions P(Secret | Male) - P(Secret | Female) = 0.0185. We can reasonably conclude that P(Secret | Gender) approximately equals P(Secret), meaning gender and secret identity status appear approximately independent in this dataset. We can conclude that knowing a character's gender gives little information about whether they have a secret identity.
